---
Lab 6 Answers File
---
This lab for BIOS 26120 (Introduction to Bioinformatics) involves building a multiple sequence alignment that can be readily applied to sequence data read in from a FASTA file.
---

```{r, include = F}
source("pairAlign.R")
sequences <- c("MVTCHQQQ", "MVRCH", "MVTCH", "QQQMVTCH", "MVTQQQCH")
buildMSA <- function(patterns, subjects, center) {
    MSA <- rbind(patterns[[1]], subjects[[1]])
    for(i in 2:length(patterns)) {
        j = 1 #index in new row
        k = 1 #index in alignment of center sequence to sequence i
        m = 1 #column index of MSA
        maxLength = ncol(MSA) + length(patterns[[i]])
        newRow = character(maxLength)
        while(k <= length(patterns[[i]]) && m <= ncol(MSA)) {
            if(patterns[[i]][k] == MSA[1, m]) {
                newRow[j] <- subjects[[i]][k]
                j <- j + 1
                k <- k + 1
                m <- m + 1
            } else if(MSA[1, m] == "-") {
                newRow[j] <- "-"
                j <- j + 1
                m <- m + 1
            } else if(patterns[[i]][k] == "-") {
                if(m == 1) {
                    MSA <- cbind(rep("-", nrow(MSA)), MSA)
                } else {
                    MSA <- cbind(MSA[, 1:(m-1)], rep("-", nrow(MSA)), MSA[, m:ncol(MSA)])
                }
                newRow[j] <- subjects[[i]][k]
                m <- m + 1
                j <- j + 1
                k <- k + 1
            }
        }
        while(k <= length(patterns[[i]])) {
            MSA <- cbind(MSA, rep("-", nrow(MSA)))
            newRow[j] <- subjects[[i]][k]
            k <- k + 1
            j <- j + 1
        }
        while(m <= ncol(MSA)) {
            newRow[j] <- "-"
            m <- m + 1
            j <- j + 1
        }
        newRow <- newRow[1:(m - 1)]
        MSA <- rbind(MSA, newRow)
    }
    rownames(MSA) <- c("    Center:", paste0("Sequence ", 1:(nrow(MSA)), ":")[-center])
    colnames(MSA) <- 1:ncol(MSA)
    return(MSA)
}
```


The buildMSA function basically combines all of the center seq’s paired alignments into one global alignment. It intakes 'patterns', a list of char vectors that each contain the version of the center sequence that comes from the pairwise alignment between the center sequence and each of the other sequences; 'subjects', a list of char vectors that each contain the alignment of the rest of the sequences with the center sequence; and 'center', the integer index of the sequence in the list that is the center. 


```{r}
#use example sequences to test alignment
sequences <- c("MVTCHQQQ", "MVRCH", "MVTCH", "QQQMVTCH", "MVTQQQCH")

n <- length(sequences)
alignmentPatterns <- list()
alignmentSubjects <- list()


for(i in 1:(n - 1)) { #for each sequence except the last one
    for(j in (i+1):n) { #for each j larger than i in sequences
        aligned <- pairAlign(sequences[i], sequences[j], substitutionMatrix = "BLOSUM62", gapOpening = -10, gapExtension=-2, type="global") #use pairAlign() here to find the pattern and subject
        alignmentPatterns <-  cbind(alignmentPatterns,aligned[1]) #store the alignment sequence of the first protein in alignmentPatterns
        alignmentSubjects <- cbind(alignmentSubjects,aligned[2]) #store the alignment sequence of the second protein in alignmentSubjects
       
    }
}
alignmentPatterns
alignmentSubjects

```


```{r}
#create matrix of distances between sequences
n <- length(sequences)

distVector <- numeric(length(alignmentPatterns)) #create distVector, a vector that's the length of alignmentPatterns
for (i in 1:length(alignmentSubjects)) {
  pattern <-strsplit(alignmentPatterns[[i]], "")[[1]]
  subject <- strsplit(alignmentSubjects[[i]], "")[[1]]
  distance <- pattern != subject
  distVector[i] <-  sum(distance) #split the two sequences to be aligned into multiple strings of amino acid characters, and count the number of mismatches between them
}
distVector #great, now you have your vector with the distances inside it

#use code from above to make a matrix that looks like exMatrix, but with your distVector instead

distMatrix <- matrix(0, nrow = n, ncol = n) #initalize with zeros
distMatrix[lower.tri(distMatrix)] <- distVector #populate lower triangle with vector
distMatrix <- distMatrix + t(distMatrix) #sum with own transposition to mirror values across diagonal
distMatrix
```


```{r}

#find sequence that is the center (with greatest distMatrix sum)- other seqs will be aligned to it

for (i in 1:n){ #Iterate through length of matrix
  maximum <- max(colSums(distMatrix)) #calculate max of column sums
  if (sum(distMatrix[,i]) == maximum){ #if that column gives the max colsum, print index
    print (i)
  }
}

#The center sequence is number 4, QQQMVTCH.

```


<span style="color:green"> 3.7 \span
```{r}
n <- length(sequences)
a <- length(alignmentPatterns)

#create a list to store our patterns and our subjects
PATTERNS <- list()
SUBJECTS <- list()
#create our 2 matrices and fill them with "-"s
subpat = matrix("-", nrow = n, ncol = n)
patsub = matrix("-", nrow = n, ncol = n)


#store elements in alignmentPatterns, a list, in alignPatternsV, a vector
alignmentPatternsV <- c()
for (i in 1:a){
  alignmentPatternsV[i] <- alignmentPatterns[[i]]
} 

#store elements in alignmentSubjects, a list, in alignSubjectsV, a vector
alignmentSubjectsV <- c()
for (i in 1:a){
  alignmentSubjectsV[i] <- alignmentSubjects[[i]]
}


#fill lower triangle of subpat with alignmentPatternsV
  subpat[lower.tri(subpat)] <- alignmentPatternsV

 #fill lower triangle of patsub with alignmentSubjectsV
  patsub[lower.tri(subpat)] <- alignmentSubjectsV

 #fill upper triangle of subpat with transposed lower triangle of patsub
  subpat[upper.tri(subpat)] <- t(patsub)[upper.tri(patsub)]

  #fill upper triangle of patsub with transposed lower triangle of subpat
  patsub[upper.tri(patsub)] <- t(subpat)[upper.tri(subpat)]


#subpat now contains subjects as rows and patterns as columns
#patsub now contains patterns as rows and subjects as columns

  
exSums <- colSums(distMatrix) #find column sums of matrix
exCandidates <- which(exSums == min(exSums)) #find which columns have minimum sum
exCenter <- exCandidates[nnet::which.is.max(nchar(sequences[exCandidates]))] #if multiple, break ties with nnet::which.is.max() to determine the one with longest name
#exCenter #index of center color


#store patterns, split into single character strings, in PATTERNS
PATTERNS <- strsplit(subpat[,exCenter][subpat[,exCenter] != "-"], "")

#store subjects, split into single character strings, in SUBJECTS
SUBJECTS <- strsplit(patsub[,exCenter][patsub[,exCenter] != "-"], "") 

print(PATTERNS)
print(SUBJECTS)

buildMSA(PATTERNS, SUBJECTS, exCenter)
```


<span style="color:green"> 3.8 \span
```{r}
#write function to implement centerStar algorithm and apply it to globins sequences (contains globin proteins from different species: human alpha and beta hemoglobin, myoglobin from physeter catodon, globin from petromyzon marinus, globin from chironomus piger, and hemoglobin from lupinus luteus)

centerStar <- function(sequences, substitutionMatrix, gapOpening, gapExtension) {
    n <- length(sequences)
    
    #find alignment patterns and subjects
    n <- length(sequences)
    alignmentPatterns <- list()
    alignmentSubjects <- list()
    
    for(i in 1:(n - 1)) { #for each sequence except the last one
        for(j in (i+1):n) { #for each j larger than i in sequences
            aligned <- pairAlign(sequences[i], sequences[j], substitutionMatrix = "BLOSUM62", gapOpening = gapOpening, gapExtension= gapExtension, type="global")
            alignmentPatterns <-  cbind(alignmentPatterns,aligned[1]) #store the alignment sequence of the first protein in alignmentPatterns
            alignmentSubjects <- cbind(alignmentSubjects,aligned[2]) #store the alignment sequence of the second protein in alignmentSubjects
          }
    }
    
    a <- length(alignmentPatterns)
    
    #create a list to store our patterns and our subjects
    PATTERNS <- list()
    SUBJECTS <- list()
    #create our 2 matrices and fill them with "-"s
    subpat = matrix("-", nrow = n, ncol = n)
    patsub = matrix("-", nrow = n, ncol = n)
    
    #store elements in alignmentPatterns, a list, in alignPatternsV, a vector
    alignmentPatternsV <- c()
    for (i in 1:a){
      alignmentPatternsV[i] <- alignmentPatterns[[i]]
    } 
    #store elements in alignmentSubjects, a list, in alignSubjectsV, a vector
    alignmentSubjectsV <- c()
    for (i in 1:a){
      alignmentSubjectsV[i] <- alignmentSubjects[[i]]
    }
    #fill lower triangle of subpat with alignmentPatternsV
      subpat[lower.tri(subpat)] <- alignmentPatternsV
     #fill lower triangle of patsub with alignmentSubjectsV
      patsub[lower.tri(subpat)] <- alignmentSubjectsV
     #fill upper triangle of subpat with transposed lower triangle of patsub
      subpat[upper.tri(subpat)] <- t(patsub)[upper.tri(patsub)]
      #fill upper triangle of patsub with transposed lower triangle of subpat
      patsub[upper.tri(patsub)] <- t(subpat)[upper.tri(subpat)]
      
      
      exSums <- colSums(distMatrix) #find column sums of matrix
      exCandidates <- which(exSums == min(exSums)) #find which columns have minimum sum
      exCenter <- exCandidates[nnet::which.is.max(nchar(sequences[exCandidates]))] #if multiple, break ties with nnet::which.is.max() to determine the one with longest name
      exCenter #index in paints vector of center color
      
      
      #store patterns, split into single character strings, in PATTERNS
      PATTERNS <- strsplit(subpat[,exCenter][subpat[,exCenter] != "-"], "")
      
      #store subjects, split into single character strings, in SUBJECTS
      SUBJECTS <- strsplit(patsub[,exCenter][patsub[,exCenter] != "-"], "") 
      
      #print(PATTERNS)
      #print(SUBJECTS)

    return(buildMSA(PATTERNS, SUBJECTS, exCenter))
}

globins <- unlist(seqinr::read.fasta("globins.fasta", seqtype = "AA", as.string = T))
centerStar(globins, "BLOSUM62", -10, -2)
```

